A crop modelling strategy to improve cacao quality and productivity

### 2021-11-23 paperModel/paperModel_2.tex
submitted

%=================================================================
\documentclass[gene,journal,article,submit,moreauthors,pdftex]{Definitions/mdpi} 
\usepackage{natbib}
\usepackage[colorinlistoftodos]{todonotes}
\usepackage[dvipsnames]{xcolor}
\newcounter{mycomment}
\newcommand{\mycomment}[2][]{%
% initials of the author (optional) + note in the margin
\refstepcounter{mycomment}%
{%
\setstretch{0.7}% spacing
\todo[color={red!100!green!33},size=\small]{%
\textbf{Comment [\uppercase{#1}\themycomment]:}~#2}%
}}
% For posting an early version of this manuscript as a preprint, you may use "preprints" as the journal and change "submit" to "accept". The document class line would be, e.g., \documentclass[preprints,article,accept,moreauthors,pdftex]{mdpi}. This is especially recommended for submission to arXiv, where line numbers should be removed before posting. For preprints.org, the editorial staff will make this change immediately prior to posting.

@@

%=================================================================
% Full title of the paper (Capitalized)
\Title{A modelling strategy to improve cacao quality and productivity }

@@

% Affiliations / Addresses (Add [1] after \address if there is only one affiliation.)
\address{$^{1}$ \quad The John Bingham Laboratory, NIAB, 93 Lawrence Weaver Road, Cambridge CB3 0LE, UK 1\\$^{2}$ \quad Grupo BIOS, Centro de Bioinformática y Biología Computacional de Colombia - BIOS, Manizales, Caldas, Colombia, South America.\\ $^{3}$ \quad Federación Nacional de Cafeteros FEDECACAO, Colombia, South America. }%; e-mail@e-mail.com}

% Contact information of the corresponding author
\corres{Correspondence: angela.romerovergel@niab.com} %; Tel.: (optional; include country code; if there are multiple corresponding authors, add author initials) +xx-xxxx-xxx-xxxx (F.L.)}

@@

% Abstract (Do not insert blank lines, i.e. \\) 
\abstract{Cacao production systems in Latin America {\color{red}have } a high importance over social and economic development, facing  the fight against hunger and poverty. Although Colombian cacao has the potential to be in the high value markets for fine flavour, {\color{red} the lack of expert support as well as the use of traditional, and oftentimes suboptimal, technologies makes cocoa production negligibly. For example, traditionally cacao harvest takes place at 5 or 6 months after flowering, other environmental parameters that have more association with pod maturation speed are not taken into account.  Cocoa fruits development can be considered as the result of a number of physiological and morphological processes that can be described by mathematical relationships even under uncontrolled environments. In this context, crop models are useful tools to simulate and predict crop development over time and under multiple environmental conditions. Since, harvesting at the right time can yield high quality cacao, we parametrised a crop model to predict the best time for harvest cocoa fruits in Colombia. Our aim was to develop a practical tool that supports cacao farmers in the production of high quality cacao.} When comparing simulated and observed data, our results showed an RRMSE of 7.2\% for the yield prediction, while the simulated harvest date varied between +/- 2 to 20 days depending on the temperature variations of the year between regions. {\color{red} This crop model contributed to understand and predict the phenology of cacao fruits for two key cacao varieties ICS95 y CCN51.}
}

% Keywords
\keyword{ICS95; CCN51; thermal time, flowering date }

@@

Cocoa ( \textit{Theobroma cacao }L. ) is an important worldwide perennial tropical crop endemic to the South American rainforests \citep{zuidema2005, motamayor2002, argout2011, Rodriguez2019}. Cacao plant is a member of the Malvaceae (formerly Sterculiaceae)  botanical family such as  cotton \textit{ Gossypium hirstium} \citep{Nix2017cotton}. Coton has been is modeled in SIMPLE model \citep{Zao2019simple}. Cocoa is grown for its fruits, known as cacao pods \citep{ Niemenak2010, suarez2021}. Only the 5\% of the world cocoa yield is destined for Fine-cocoa production due to the low productivity through  the traditional crop management \citep{argout2011}.  In Colombia, cocoa  is  traditionally  consumed  as  a  beverage. It is one of the crops promoted by the Colombian government in the social and agricultural development  programs aimed at favouring peace in post-conflict regions \citep{Rodriguez2019, Abbott2019}. This crop is grown by approximately 52.000 families \citep{Gutierrez2020} and 98\% of production being carried out by small and medium-sized producers \citep{Garcia2014, Escobar2020}. Colombia registered an increase of 3.750 tons in production in 2020 compared to the previous year \citep{lamos2020}.

Although Colombian cocoa has the potential to be in the high value markets for fine flavour \citep{Escobar2020}, {\color{red} the lack of expert support as well as the use of traditional, and often times suboptimal, technologies makes cocoa production negligibly. For example, the farmer practice of empirically harvesting from 5 to 6 months or {\color{olive} 180 days } after flowering date (DAF), produced a mix of quality of cocoa beans that are then be fermented. This practice produces heterogeneous characteristics between each fermentation batch which potentially diminish the quality of the end cocoa product} \citep{Escobar2021}. To identify the best moment to harvest is important to consider physiological responses affected by climate variables such as rain, solar radiation and wind.  Thus,  for cocoa in Colombia, physiological simulation models may be valuable to identify the best moment to harvest cocoa considering variable weather conditions, soil types and cultivar specifications.

Crop models represent a quantitative assumption of plant growth depending on sunlight interception efficiency values and climate data supported by a large amount of empirical and ground data. \citep{Reynolds2018}. Physiological crop models have shown to be very useful tool for provided agronomical advices and improvements of the cropping systems of annual crops mainly. Recently crop modelling studies are focusing on  perennial crops  production \citep{zuidema2005, Zao2019simple, Bai2020, Romero2021}. However, the information reported  is be less than for annual crops due to the lack of field data available , relatively high research costs and the difficulties of accumulated errors in long-term simulations \citep{zuidema2005}. For cacao there  approaches  to predict yield mainly  using algorithms of machine learning \citep{lamos2020} and just one mechanistic model simulates physiological cocoa performance  "SUCROS-cocoa" \citep{zuidema2005}. This crop model calculates light interception, photosynthesis, maintenance respiration,
evapotranspiration, biomass production and cocoa yield. It can be parametrised having data on cocoa physiology and morphology \citep{zuidema2005}. However,  there is not specific cocoa physiology data available from small and medium-sized producers. Thus, we adapted the simple generic crop model (SIMPLE) that could be easily modified for any crop to simulate development, crop growth and yield using few parameters such as weather and cultivar specification \citep{Zao2019simple}.

In this paper, we present a physiological parametrisation of SIMPLE crop model for cocoa to predict best harvest time and overall yield. {\color{red} We parameterised the SIMPLE crop model \citep{Zao2019simple} because it had already been successfully fitted to other tree crops in south America. The model simulates crop development, growth and yield, and predicts the maturation day when the fruit is likely to be ready for harvest. Our aim was to develop a practical tool that supports cacao farmers in the production of high quality cacao. Thus, making Colombian cacao more competitive in the Fine-cocoa market.}

@@

Usually, the phenological stages of a cacao tree are divided in two main phases: vegetative and reproductive. Defining the reproductive phase (fig. \ref{fig:pheno}, a) as described by the floral phenology from the date of inflorescence emergence (BBCH scale 5) (fig. \ref{fig:pheno},b ) to the date of ripening of fruit and seed (BBCH 8) (fig. \ref{fig:pheno},c ) \citep{Niemenak2010}. The reproductive phase in the Andean region of Colombia is cyclically fulfilled during two annual cycles which goes through the following phases: inflorescence emergence, flowering, pollination, fruit development and harvest. Therefore, for modelling parameterisation the crop cycle of cocoa as perennial plant does not start at the plantation date such as annual crops systems. Instead, the start point of the cocoa crop cycle is the inflorescence emergence date (fig. \ref{fig:pheno},b). Consequently , the growth period of the fruit can vary from  110 to 150  daa (days after anthesis) \citep{lopez2018} when cacao fruits reach the physiological maturity, but it can be harvested at 170 days daa \citep{Niemenak2010} for quality purposes.

{\color{red} Cocoa is a cauliflorous plant, which means that flowers grow on the trunk and branches. Cocoa trees usually produce up to 10000 flowers per tree each year, 50 \%  of them do not develop into ripe fruits (Personal communication with Fedecacao). Flower development takes approximately 30 days across 12 micro-stages, from} meristem development (stages 1 to 6) to the fully developed flower (Stages 7 to 12) \citep{swanson2005, lopez2018} when it is ready to be pollinated. The opening of flowers or anthesis  occurs over a 12-hour period during the night and it is synchronised between the groups of mature flowers \citep{Niemenak2010}. However, the live of a flower can last approximately 1 day after the opening falling form the trunk if it is unfertilised \citep{cheesman1927, Niemenak2010}. {\color{red} Subsequentially, after anthesis, fruits growth for approximately 150 days until the maturation, mucilage}. Therefore, the complete maturation process of the fruit, from the pollination to fully mature fruit, takes 160- 210 days \citep{berry1994}. The accumulation of lipids, storage proteins and anthocyanin start about 85 days after pollination when fruits have an active metabolism and seeds moisture content decreases up to 30\% \citep{Lehrian1980, Niemenak2010}. During this phase the quality of cocoa seeds is defined.

\begin{figure}[h!]
	\centering
	\includegraphics[scale=0.4]{images/phenology.png}\\
	\caption{\footnotesize {Phenology of cocoa in Colombia for crop modelling.\\}} 
	\footnotesize{Source: Taken from from Dreamstime.com, phys.org \citep{toledo2021} and \cite{lopez2018}}
	\label{fig:pheno}
\end{figure}

@@

{\color{olive} Input variables required to run the cocoa model included the flowering date and daily weather of solar radiation (SRAD), maximum and minimum temperature (TMAX, TMIN) and rain. Therefore, weather data as csv file was  downloaded from the POWER Data Access Viewer \citep{nasapower} from January 01 of 2018 to December 31 of 2020 for the five locations in Colombia (Fig.\ref{fig:yield}, b). The csv file had to be transformed to .WHT file using R 1.4 version \citep{Rstudio2020}.

Field data for this study was provide by Fedecacao  (National Cacao Producers Federation). Reports contained information of 23 flowering dates from 12-July-2019 to 23-June-2020. Each of these dates had  their corresponding date of harvest after at exactly 180 DAF,  the age of the trees, plant density (trees ha$^{-1}$), yield  (dry beans kg ha$^{-1}$) and number of fruits harvested per hectare. Thus, 23 flowering dates for five farms gave us 115 samples in total.}

{\color{olive} Five farms in different regions were tested : Saravena (Arauca), Rionegro (Santander), Cali (Valle del Cauca), Apartado (Antioquia) and Manizalez (Caldas). These were considered because had the data available of production in the reports of  Fedecacao. Yield depend on the successfully development of flowers to form ripe pods. } {\color{red}  According to personal communication from farmers, the highest flowering season occurs in September and January suggesting that harvest occurs in March and July. However, the data collected from the farms suggested that flowering and pod production were not constant for all the locations. For example, pod harvest in Caldas farm increased in May and from October to December, and decreased from January to March.  Meanwhile, the Arauca and Apartado farms reported the highest yield in the months January, July, November and December. Santander farm had picks of production in March, May and September (Fig.\ref{fig:yield}, a).} {\color{olive} However,  cacao is cultivated in 30 regions of Colombian which accounts for approximately about 147,000 ha \citep{Meza2021}. Therefore, the national production of cocoa beans in 2020 was  63.416 ton according  to  information  published by Fedecacao (\citep{Fede2021}). The highest yield was reposted by Santander region with 26,315,  following in descending order by Antioquia with 5.974 tons, Arauca with 5.082 tons, Caldas with 1.343 tons and Valle del Cauca with 339 tons.}

\begin{figure}[h]
	\centering
	\includegraphics[scale=0.3]{images/map.png}\\
	\caption{\footnotesize {Cocoa production  of five farms in different regions.\\}}
	\label{fig:yield}
	{\footnotesize (a) Production per month (2019 - 2020) of five farms. (b) Map showing the regions where farms are located.}
\end{figure}
\newpage

@@

{\color{olive} The cumulative sum of daily temperature from a reference day 0 is defined as \emph{Thermal time} and  its units of measurement are in days degrees (days $^\circ$C). That starter point of 0 days $^\circ$C generally is the planting date \citep{Ritchie1991} but for cocoa crop simulation model it is the flowering date.  Thermal time of a cultivated plant may consider the base temperature (T$_{b}$), which is the minimum temperature required by cocoa plant to grow. T$_{b}$ an vary between cultivars \citep{Slafer1995, Daymond2008}. For cocoa the vegetative growth T$_{b}$ has been reported between 18.6 and 20.8 $^\circ$C \citep{Daymond2008}. Nevertheless, the pod growth  has a lower T$_{b}$ which range between 9 and 12.9 $^\circ$C \citep{Daymond2008, lahive2019}. We calculated the cocoa thermal time with a pod growth T$_{b}$ of 10 $^\circ$C because it is the absolute minimum temperature for cocoa growing in South America reported by \cite{Erneholm1948}, in \citep{lahive2019}.

The thermal time required for the crop model was characterised for each location starting from flowering date (0 days $^\circ$C) to harvest date (180 after flowering = 6 months), as farmers used to harvest by calendar days. }  Thus, the cocoa model predicts the maturation day to harvest pods. It can vary depending on temperature variations.  Thermal time was calculated using the equation \ref{equ:tem}. Where tt is the cumulative sum of the daily temperature (T$_{i}$) and T$_{b}$ for cocoa is 10$^\circ$C.

@@

Calibration of crop models are conducted typically for particular cultivars and require site specific inputs of weather \citep{Crout20142}. The procedure for the SIMPLE model \citep{Zao2019simple} calibration  was a sequential process of modifying physiological variables specific for cocoa in the inputs files and  adding the appropriate files of weather for each region tested. In the SIMPLE model has seven input files where new cocoa crop data should be provided. Files in the list below can be edited to define the features of the new cultivars or experiments. Cocoa crop has not been simulated with SIMPLE model, hence consider previous cocoa studies, values such as leaf area index (LAI) \citep{Agele2016} and Harvest index (HI) \citep{Quintana2015} where modified. Radiation Use Efficiency (RUE) was calibrate according to the perennial crops  (banana and cotton) that had been calibrated previously in SIMPLE model \citep{Zao2019simple} with a RUE of 0.8 and 0.85 respectively. As cocoa trees are under shadow the RUE was lower with values between 0.7 and 0.5 g MJ$^{\mathbf{-1}}$ m$^{2}$ (table \ref{tab:reparam}).
Once the physiological parameters were calibrated to the  simulated yields for cocoa were reasonably close to the observed yield. The 23 flowering from Fedecacao reports were introduced in the treatment file, running the program and saving the results.

@@

{\color{olive}This cocoa model has  were three parameters which vary by region (table \ref{tab:reparam}): The thermal time required for harvest after the flowering date (Tsum) , the Radiation Use Efficiency (RUE) and yield observed on field. Physiological parameters in table \ref{tab:Treaparam} are  common  for all the regions studied. These parameters were calibrate for cultivars ICS95 and CCN51 considering a range of time of 200 DAF from flowering date to harvest day, even thought farmers collect the pod at 180 DAF. Heat and water stress parameters were not considered. }

@@

\begin{table}[h!]	
	\caption {\footnotesize {Parameter values used to run SIMPLEcocoa model.}}
	\centering
	\label{tab:Treaparam} 
	\begin{small}
		\begin{tabular}{{l l l}}
			\hline
			{\bf File }&{\bf Variable name }&{\bf Value}\\
			\hline
			&SoilName & Loamy sand4\\
			&InitialFsolar & 0.01\\
			Treatment&Weather & KOKO (.WTH file name)\\
			&CO$_{2}$ & 400 ppm\\
			&SowingDate &Flowering date\\
			\hline
			&Crop cycle DAP & 200 days\\
			&LAI & 1.8 \\
			Observation&FSolar& 0.70\\
			&Biomass & 40kg dry mass per plant\\
			\hline
			&Harvest index & 0.3\\
			Cultivar&150A & 680 $^\circ$C day \\
			&150B & 680 $^\circ$C day \\
			\hline
			&Tbase & 10$^\circ$C\\
			&Topti & 26$^\circ$C \\
			Species&MaxT & 35$^\circ$C \\
			&ExtremeT & 40$^\circ$C  \\
			&CO$_{2}$RUE & 0.09$^\circ$C  \\			
			&S-water & 0 ARID index \\
			\hline			
		\end{tabular} \\ 
	\end{small}
	{\footnotesize S-water is associated drought stress evaluations ranging from 0 (no water shortage) to 1 (extreme water shortage) \cite{Zao2019simple} }
\end{table}
\newpage

@@

The cocoa model performance was evaluated by comparing simulated values cocoa yield with those reported by Fedecacao from cocoa plantations, using the statistical indice of relative root mean square error (RRMSE) (Equ. \ref{equ:RMSE}) \citep{Zao2019simple, Bai2020}.

\begin{equation}
RMSE= \sqrt{\frac{1}{n}  \sum_{i=1}^{n} (Y_{i}-X_{i})^{2} } 
\label{equ:RMSE}
\end{equation}

\section{RESULT}

\subsection{Weather Conditions Over Flowering Time }

>>>>

%=================================================================
\documentclass[gene,journal,article,submit,moreauthors,pdftex]{Definitions/mdpi} 
\usepackage{natbib}
\usepackage[colorinlistoftodos]{todonotes}
%\usepackage[dvipsnames]{xcolor}
\newcounter{mycomment}
\newcommand{\mycomment}[2][]{%
% initials of the author (optional) + note in the margin
\refstepcounter{mycomment}%
{%
\setstretch{0.7}% spacing
\todo[color={red!100!green!33},size=\small]{%
\textbf{Comment [\uppercase{#1}\themycomment]:}~#2}%
}}
% For posting an early version of this manuscript as a preprint, you may use "preprints" as the journal and change "submit" to "accept". The document class line would be, e.g., \documentclass[preprints,article,accept,moreauthors,pdftex]{mdpi}. This is especially recommended for submission to arXiv, where line numbers should be removed before posting. For preprints.org, the editorial staff will make this change immediately prior to posting.

@@

%=================================================================
% Full title of the paper (Capitalized)
\Title{A crop modelling strategy to improve cacao quality and productivity }

@@

% Affiliations / Addresses (Add [1] after \address if there is only one affiliation.)
\address{$^{1}$ \quad The John Bingham Laboratory, NIAB, 93 Lawrence Weaver Road, Cambridge CB3 0LE, UK \\$^{2}$ \quad Grupo BIOS, Centro de Bioinformática y Biología Computacional de Colombia - BIOS, Manizales, Caldas, Colombia, South America.\\ $^{3}$ \quad Federación Nacional de Cafeteros FEDECACAO, Colombia, South America. }%; e-mail@e-mail.com}

% Contact information of the corresponding author
\corres{Correspondence: ing.aromerov@gmail.com; anyela.camargorodriguez@niab.com} %; Tel.: (optional; include country code; if there are multiple corresponding authors, add author initials) +xx-xxxx-xxx-xxxx (F.L.)}

@@

% Abstract (Do not insert blank lines, i.e. \\)

\abstract{Cacao production systems in Latin America have a high importance over social and economic development, facing  the fight against hunger and poverty. Although Colombian cacao has the potential to be in the high value markets for fine flavour, the lack of expert support as well as the use of traditional, and often times suboptimal, technologies makes cocoa production negligibly. For example, traditionally cacao harvest takes place at 5 or 6 months after flowering, other environmental parameters that have more association with pod maturation speed are not taken into account.  Cacao fruits development can be considered as the result of a number of physiological and morphological processes that can be described by mathematical relationships even under uncontrolled environments. In this context, crop models are useful tools to simulate and predict crop development over time and under multiple environmental conditions. Since, harvesting at the right time can yield high quality cacao, we parametrised a crop model to predict the best time for harvest cocoa fruits in Colombia. Our aim was to develop a practical tool that supports cacao farmers in the production of high quality cacao. When comparing simulated and observed data, our results showed an RRMSE of 7.2\% for the yield prediction, while the simulated harvest date varied between +/- 2 to 20 days depending on the temperature variations of the year between regions. This crop model contributed to understand and predict the phenology of cacao fruits for two key cultivars ICS95 y CCN51.}

% Keywords
\keyword{thermal time, flowering date (FD), harvest prediction, biomass, light interception }

@@

Cacao ( \textit{Theobroma cacao }L. ) is an important worldwide perennial tropical crop endemic to the South American rainforests \citep{zuidema2005, motamayor2002, argout2011, Rodriguez2019}. Cacao plant is a member of the Malvaceae (formerly Sterculiaceae)  botanical family such as  cotton \textit{ Gossypium hirstium} \citep{Nix2017cotton}. Cotton has been is modeled in SIMPLE model \citep{Zao2019simple}. Cacao is grown for its fruits, known as cacao pods \citep{ Niemenak2010, suarez2021}. Only the 5\% of the world cocoa yield is destined for Fine-cocoa production due to the low productivity through  the traditional crop management \citep{argout2011}.  In Colombia, cocoa  is  traditionally  consumed  as  a  beverage. It is one of the crops promoted by the Colombian government in the social and agricultural development  programs aimed at favouring peace in post-conflict regions \citep{Rodriguez2019, Abbott2019}. This crop is grown by approximately 52.000 families \citep{Gutierrez2020} and 98\% of production being carried out by small and medium-sized producers \citep{Garcia2014, Escobar2020}. Colombia registered an increase of 3.750 tons in production in 2020 compared to the previous year \citep{lamos2020}.

Although Colombian cocoa has the potential to be in the high value markets for fine flavour \citep{Escobar2020}, the lack of expert support as well as the use of traditional, and often times suboptimal, technologies makes cocoa production negligibly. For example, the farmer practice of empirically harvesting from 5 to 6 months or 180 Days After Flowering  (DAF) date, produced a mix of quality of cacao beans that are then fermented. This practice produces heterogeneous characteristics between each fermentation batch which potentially diminish the quality of the final cocoa product \citep{Escobar2021}. To identify the best moment to harvest is important to consider physiological responses affected by climate variables such as rain, solar radiation and wind.  Therefore,  for cacao production in Colombia, physiological simulation models may be valuable to identify the best moment to harvest cocoa fruits considering variable weather conditions, soil types and cultivar specifications.

Crop models represent a quantitative assumption of plant growth depending on sunlight interception efficiency values and climate data supported by a large amount of empirical and ground data. \citep{Reynolds2018}. Physiological crop models have shown to be very useful tool for provided agronomical advices and improvements of the cropping systems of annual crops mainly. Recently crop modelling studies are focusing on  perennial crops  production \citep{zuidema2005, Zao2019simple, Bai2020, Romero2021}. However, the information reported  is be less than for annual crops due to the lack of field data available , relatively high research costs and the difficulties of accumulated errors in long-term simulations \citep{zuidema2005}. For cacao there are  approaches  to predict yield mainly  using algorithms of machine learning \citep{lamos2020} and just one mechanistic model simulates physiological cocoa performance  "SUCROS-cocoa" \citep{zuidema2005}. This crop model calculates light interception, photosynthesis, maintenance respiration,
evapotranspiration, biomass production and cocoa yield. It can be parametrised having data on cocoa physiology and morphology \citep{zuidema2005}. However,  there is not specific cocoa physiology data available from small and medium-sized producers. Thus, we adapted the simple generic crop model (SIMPLE) that could be easily modified for any crop to simulate development, crop growth and yield using few parameters such as weather and cultivar specification \citep{Zao2019simple}.

In this paper, we present a physiological parametrisation of the SIMPLE crop model \citep{Zao2019simple} for cocoa to predict best harvest time and overall yield in the framework of KOCOLATL project, which is a collaborative work between Colombian partners (researches, farmers and organizations such as Fedecacao and BIOS) and researchers in NIAB from United Kingdom. We parametrised this crop model  because it had already been successfully fitted to other tree crops in south America. The model simulates crop development, growth and yield, and predicts the maturation day when the fruit is likely to be ready for harvest. Our aim was to develop a practical tool that supports cacao farmers in the production of high quality cocoa. Thus, making Colombian cacao more competitive in the Fine-cocoa market.

@@

The phenological stages of a cacao tree are divided in two main phases, vegetative and reproductive. The reproductive phase (fig. \ref{fig:pheno}, a) is represented by the floral phenology which goes from the date of inflorescence emergence (BBCH scale 5) (fig. \ref{fig:pheno},b ) to the date of ripening of fruit and seed (BBCH 8) (fig. \ref{fig:pheno},c ) \citep{Niemenak2010}. The reproductive phase is cyclically across two annual cycles which go through the following phases: inflorescence emergence, flowering, pollination, fruit development and harvest. Therefore, to model the cacao crop cycle as perennial plant, we regarded the the start point of the cocoa crop cycle as the inflorescence emergence date, as opposed to the sowing/emergence date usually used to model annual crops (fig. \ref{fig:pheno},b). Therefore, the growth period of the fruit could vary from 110 to 150 daa (days after anthesis) \citep{lopez2018} when cacao fruits reach the physiological maturity, but it can be harvested at 170 days daa \citep{Niemenak2010} for quality purposes.

Since cacao is a cauliflorous plant, flowers grow on trunk and branches. Cacao trees usually produce up to 10000 flowers per tree each year, 50 \%  of which do not develop into ripe fruits (Personal communication with Fedecacao). Flower development takes approximately 30 days across 12 micro-stages, from meristem growth (stages 1 to 6) to the fully developed flower (Stages 7 to 12) \citep{swanson2005, lopez2018} when it is ready to be pollinated. The opening of flowers or anthesis  occurs over a 12-hour period during the night and it is synchronised between the groups of mature flowers \citep{Niemenak2010}. However, the life of a flower can last approximately 1 day after the opening, when it falls form the trunk if it is unfertilised \citep{cheesman1927, Niemenak2010}. Subsequentially, after anthesis, fruits growth for approximately 150 days until the maturation, mucilage. Therefore, the complete maturation process of the fruit, from the pollination to fully mature fruit, takes 160- 210 days \citep{berry1994}. The accumulation of lipids, storage proteins and anthocyanin start about 85 days after pollination when fruits have an active metabolism and seeds moisture content decreases up to 30\% \citep{Lehrian1980, Niemenak2010}. During this phase the quality of cocoa seeds is defined.

\begin{figure}[h!]
	\centering
	\includegraphics[scale=0.15]{images/phenology.png}\\
	\caption{\footnotesize {Phenology of cocoa in Colombia for crop modelling.\\}} 
	\footnotesize{Source: Taken from Dreamstime.com, phys.org \citep{toledo2021} and \cite{lopez2018}}
	\label{fig:pheno}
\end{figure}

To calibrate and evaluate the cacao model, 23 field data samples were collected by Fedecacao, from five farms located in the geographic regions of Saravena (Arauca), Rionegro (Santander), Cali (Valle del Cauca), Apartado (Antioquia) and Manizalez (Caldas) (Fig.\ref{fig:yield}, a). Each field sample reported an observed flowering date and corresponding yield, for a given tree and month of the year. Field samples reported sights of 23 flowering dates from 12-July-2019 to 23-June-2020. Each of these dates had their corresponding date of harvest after at exactly 180 DAF,  the age of the trees, plant density (trees ha$^{-1}$), yield  (dry beans kg ha$^{-1}$) and number of fruits harvested per hectare. Thus, 23 flowering dates for five farms gave us 115 samples in total.

@@

In order to model the phenology of the cacao fruit, linked to collected field data, the cacao model required specific input variables such as Flowering Date (FD), Daily Solar Radiation (SRAD), Daily Maximum and Minimum Temperature (TMAX, TMIN) and Daily Precipitation, specific to a particular agroclimatic region. FD was extracted from the field data, corresponding weather data linked to geographical region was sourced from the POWER Data Access Viewer \citep{nasapower}, from January 01 of 2018 to December 31 of 2020 (Fig.\ref{fig:yield}, b). Weather data was processed using R (R 1.4 version) \citep{Rstudio2020}.

Since cacao yield depends on the successful development of flowers to form ripe pods, the current view from Colombian farmers (personal communication with Fedecacao)  is that the highest flowering season occurs in September and January which suggests that harvest occurs in March and July. These communication were given through weekly online meetings of KOCOLATL project.  However, the field data collected from Fedecacao suggested that flowering and pod production were not constant for all the locations. For example, pod harvest in the Caldas farm increased in May, and from October to December, and decreased from January to March.  The Arauca and Apartado farms reported highest yield in the months of January, July, November and December. The Santander farm had picks of production in March, May and September (Fig.\ref{fig:yield}, b).

\begin{figure}[h]
	\centering
	\includegraphics[scale=0.3]{images/map.png}\\
	\caption{\footnotesize {Cocoa production  of five farms in different regions.\\}}
	\label{fig:yield}
	{\footnotesize (a) Map showing the regions where farms are located.  (b) Production per month (2019 - 2020) of five farms.}\\
\end{figure}

@@

The cumulative sum of daily temperature from a reference day 0 is defined as \emph{Thermal time} and  its units of measurement are in days degrees (days $^\circ$C). That starter point of 0 days $^\circ$C generally is the planting date \citep{Ritchie1991} but as indicated earlier, our cacao model used FD as the starting point. Thermal time for the development of a cultivated plant may consider the base temperature (T$_{b}$), which is the minimum temperature required by cacao plants to grow. In addition, T$_{b}$ can vary between cultivars \citep{Slafer1995, Daymond2008}. For cacao the vegetative growth T$_{b}$ has been reported between 18.6 and 20.8 $^\circ$C \citep{Daymond2008}. Nevertheless, the pod growth  has a lower T$_{b}$ which range between 9 and 12.9 $^\circ$C \citep{Daymond2008, lahive2019}. We calculated the cacao thermal time with a pod growth T$_{b}$ of 10 $^\circ$C because it is the absolute minimum temperature for cocoa growing in South America reported by \cite{Erneholm1948}, in \citep{lahive2019}.

The thermal time required for the cacao crop model was characterised for each location starting from FD (0 days $^\circ$C) to harvest date (180 after flowering = 6 months), as farmers used to harvest by calendar days. Thus, the cacao model predicts the maturation day to harvest pods. It can vary depending on temperature variations.  Thermal time was calculated using the equation \ref{equ:tem}. Where tt is the cumulative sum of the daily temperature (T$_{i}$) and T$_{b}$ for cocoa is 10$^\circ$C.

@@

Calibration of any crop model is conducted typically for a particular cultivar and agroclimatic region \citep{Crout20142}. The cacao model was calibrated, for five particular regions and two cultivars, by sequentially modifying the physiological variables and then comparing the degree of similarity between observed and predicted data \citep{Crout2008chapter, Zao2019simple,Camargo2019aquacropr, Bai2020}. The original code of SIMPLE model from \citep{Zao2019simple} was modified into the cocoa model. The process of calibration stops when the distance between observed and simulated data doesn't improved any longer. The cacao model has seven input files where new cocoa crop data are provided. Files in the list below can be edited to define the features of the new cultivars or experiments. Since cacao phenology has not been simulated with the SIMPLE model, cacao physiological information such as Leaf Area Index (LAI) for shade plants \citep{Agele2016, Soltani2012, zuidema2005,Baracaldo2014} and Harvest Index (HI) \citep{Quintana2015} were extracted from the current literature and verified with Fedecacao cacao experts. Other variables such as Radiation Use Efficiency (RUE) \citep{Fletcher2013RUE,Bonhomme2000} were based on the perennial crops banana and cotton which have been calibrated previously in the SIMPLE model \citep{Zao2019simple} using RUEs of 0.8 and 0.85 for banana and cotton, respectively. As cocoa trees usually grow under shadow \citep{lahive2019}, we estimated by trial and error that lower RUE values (between 0.7 and 0.5 g MJ$^{\mathbf{-1}}$ m$^{2}$) than those utilized by \cite{Zao2019simple}, was a most ideal range  (table \ref{tab:reparam}).

@@

This cocoa model has three parameters which vary by region (table \ref{tab:reparam}): The thermal time required for harvest after the FD (Tsum) , the Radiation Use Efficiency (RUE) and yield observed on field. Physiological parameters in table \ref{tab:Treaparam} are  common  for all the regions studied. These parameters were calibrate for cultivars ICS95 and CCN51 considering a range of time of 200 DAF to harvest day, even thought farmers collect the pod at 180 DAF. Heat and water stress parameters were not considered as this study was not assessing biotic o abiotic stresses.

@@

\begin{table}[h!]	
	\caption {\footnotesize {Parameter values used to run the cacao model.}}
	\centering
	\label{tab:Treaparam} 
	\begin{small}
		\begin{tabular}{{l l l}}
			\hline
			{\bf File }&{\bf Variable name }&{\bf Value}\\
			\hline
			&SoilName & Loamy sand4\\
			&InitialFsolar & 0.01\\
			Treatment&Weather & KOKO (.WTH file name)\\
			&CO$_{2}$ & 400 ppm\\
			&SowingDate &Flowering Date (FD)\\
			\hline
			&Crop cycle DAP & 200 days\\
			&LAI & 1.8 \\
			Observation&FSolar& 0.70\\
			&Biomass & 40kg dry mass per plant\\
			\hline
			&Harvest index & 0.3\\
			Cultivar&150A & 680 $^\circ$C day \\
			&150B & 680 $^\circ$C day \\
			\hline
			&Tbase & 10$^\circ$C\\
			&Topti & 26$^\circ$C \\
			Species&MaxT & 35$^\circ$C \\
			&ExtremeT & 40$^\circ$C  \\
			&CO$_{2}$RUE & 0.09$^\circ$C  \\			
			&S-water & 0 ARID index \\
			\hline			
		\end{tabular} \\ 
	\end{small}
	{\footnotesize S-water is associated drought stress evaluations ranging from 0 (no water shortage) to 1 (extreme water shortage) \cite{Zao2019simple} }
\end{table}
\newpage

@@

The cocoa model performance was evaluated by comparing simulated values cocoa yield with those reported by Fedecacao from cocoa plantations, using the statistical indice of Relative Root Mean Square Error (RRMSE)  in the equation \ref{equ:RRMSE}, where n is the total number of observations, Y$_{i}$ corresponds to observed value from field and X$_{i}$ is the predict value from the model.\\

\begin{equation}
RRMSE= \sqrt{\frac{\frac{1}{n}  \sum_{i=1}^{n} (Y_{i}-X_{i})^{2}}{\sum_{i=1}^{n} X_{i}^{2} } } \times 100\% 
\label{equ:RRMSE}
\end{equation}

\section{Results}

\subsection{Weather Conditions Over Flowering Time }
We analysed and compared weather patterns linked our five geographical regions (Fig.\ref{fig:yield}, a) and across 2018 to 2020 . Looking at the data, the regions of Santander and Caldas had the biggest variability and the maximum of solar radiation values over 20 MJ m$^{2}$day$^{-1}$ . In contrast , the regions of Cali and Arauca presented the lowest values of PAR below 5 MJ m$^{2}$day$^{-1}$. (Fig.\ref{fig:temp}, a).  Even though, Cali and Santander had contrasting PAR conditions, these regions presented  have similar temperature during 2020 (Fig.\ref{fig:temp}, b). The temperature ranges from 16 to 28 $^\circ$C and it is relatively constant for each region. However, Arauca presented the biggest variability with hotter months during the first half of the year 2019 and 2020.  Apartado was found as the hottest region studied with a relative constant temperature of 26 $^\circ$C and Caldas as the coldest site with 16 $^\circ$C.  Precipitation in Colombia is presented in two seasons per year from February to April and from October to November, while the relative humidity remain constant over 80\% (Fig.\ref{fig:temp},c). In general, the coldest regions tested (Caldas and Santander) had the maximum values of solar radiation available for photosynthesis.
